---
title: "Regularization Methods: Ridge, Lasso, and Elastic Net"
author: "Bill Cresko"
format: 
  revealjs:
    footer: BioE_Stats_2026 - Knight Campus 
    transition: fade
    transition-speed: slow
    theme: default
    slide-number: true
    code-fold: false
    code-overflow: wrap
    code-line-numbers: true
    highlight-style: github
    embed-resources: true
    css: cresko_style.css
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(glmnet)
library(knitr)

theme_set(theme_minimal(base_size = 14))
set.seed(2026)
```

# Regularization Methods {background-color="#2c3e50"}

## From Binary Outcomes to High-Dimensional Predictors

Odds ratios help us understand individual predictor effects. But what happens when we have **many predictors** --- possibly more than our sample size? Standard regression breaks down.

**Regularization** methods (Ridge, Lasso, Elastic Net) solve this by adding penalty terms that **shrink coefficients** toward zero, preventing overfitting in high-dimensional bioengineering data like proteomics, metabolomics, or comprehensive material characterization.

## Biomaterial Properties Dataset (High-Dimensional)

::: callout-note
### Study Description

A materials science lab characterized 100 scaffold formulations using 15 measured properties (porosity, stiffness, fiber diameter, surface chemistry metrics, etc.). They want to predict a continuous biocompatibility score from these properties, but many are correlated and some may be irrelevant. With 15 predictors and only 100 observations, standard OLS regression risks overfitting.
:::

| Variable | Type | Description | Notes |
|:---|:---|:---|:---|
| `property_1`--`property_15` | Continuous | 15 scaffold characterization metrics | Standardized |
| `biocompat_score` | Continuous | Continuous biocompatibility score | Response variable |

```{r}
#| label: biomaterial-hd-load
#| eval: false
#| echo: true

# Load the high-dimensional biomaterial properties dataset
biomaterial_hd <- read_csv("data/biomaterial_properties_hd.csv")
X <- biomaterial_hd |> select(starts_with("property_")) |> as.matrix()
y <- biomaterial_hd$biocompat_score
glimpse(biomaterial_hd)
```

## Why Regularization?

Standard linear regression struggles when:

-   **Multicollinearity**: Correlated predictors lead to unstable coefficient estimates
-   **High dimensionality**: More predictors than observations (p \> n)
-   **Overfitting**: Model fits training data too closely, poor generalization

**Solution**: Add a penalty term that **shrinks coefficients toward zero**, trading a small amount of bias for a large reduction in variance.

## Comparing Regularization Methods

| Method | Penalty | Variable Selection | Handles Correlation | R Parameter |
|:---|:---|:---|:---|:---|
| **Ridge (L2)** | $\lambda \sum \beta_j^2$ | No (shrinks all) | Yes | `alpha = 0` |
| **Lasso (L1)** | $\lambda \sum |\beta_j|$ | Yes (zeros some) | Picks one of group | `alpha = 1` |
| **Elastic Net** | Both L1 + L2 | Yes | Yes | `0 < alpha < 1` |

## Ridge Regression (L2 Penalty)

::: panel-tabset
### Equation

$$\hat{\beta}_{ridge} = \arg\min_{\beta} \left[ \sum_{i=1}^{n} (y_i - x_i^T\beta)^2 + \lambda \sum_{j=1}^{p} \beta_j^2 \right]$$

### LaTeX

``` text
\hat{\beta}_{ridge} = \arg\min_{\beta} \left[ \sum_{i=1}^{n} (y_i - x_i^T\beta)^2 + \lambda \sum_{j=1}^{p} \beta_j^2 \right]
```

### Properties

-   Penalty: $\sum \beta_j^2$ (L2 norm — sum of squared coefficients)
-   Shrinks coefficients toward zero but **never exactly zero**
-   Handles multicollinearity well
-   Closed-form solution: $\hat{\beta}_{ridge} = (X^TX + \lambda I)^{-1}X^Ty$
-   As λ → 0: approaches OLS; as λ → ∞: coefficients → 0
-   Ridge is ideal when you have correlated biomaterial properties and believe all predictors contribute
:::

## Lasso Regression (L1 Penalty)

::: panel-tabset
### Equation

$$\hat{\beta}_{lasso} = \arg\min_{\beta} \left[ \sum_{i=1}^{n} (y_i - x_i^T\beta)^2 + \lambda \sum_{j=1}^{p} |\beta_j| \right]$$

### LaTeX

``` text
\hat{\beta}_{lasso} = \arg\min_{\beta} \left[ \sum_{i=1}^{n} (y_i - x_i^T\beta)^2 + \lambda \sum_{j=1}^{p} |\beta_j| \right]
```

### Properties

-   Penalty: $\sum |\beta_j|$ (L1 norm — sum of absolute values)
-   Can set coefficients **exactly to zero** → performs variable selection
-   No closed-form solution (requires optimization)
-   Produces **sparse models** — useful when many predictors are irrelevant
-   **Caution**: Among a group of correlated predictors, Lasso arbitrarily selects one and zeroes out the others
:::

## Elastic Net

Combines Ridge and Lasso penalties:

::: panel-tabset
### Equation

$$\hat{\beta}_{elastic} = \arg\min_{\beta} \left[ RSS + \lambda \left( \alpha \sum |\beta_j| + (1-\alpha) \sum \beta_j^2 \right) \right]$$

### LaTeX

``` text
\hat{\beta}_{elastic} = \arg\min_{\beta} \left[ RSS + \lambda \left( \alpha \sum |\beta_j| + (1-\alpha) \sum \beta_j^2 \right) \right]
```
:::

-   α = 0: pure Ridge; α = 1: pure Lasso
-   Variable selection like Lasso + handles correlated features like Ridge
-   Often the best default choice

::: callout-tip
### Practical Guidance

In bioengineering high-dimensional data (proteomics, material characterization, gene expression), start with **Elastic Net (α = 0.5)** as a safe default. If you need a minimal predictor set, increase α toward 1 (Lasso). If prediction accuracy matters more than variable selection, decrease α toward 0 (Ridge). Use `cv.glmnet()` to optimize both α and λ simultaneously.
:::

## Regularization in R

::: panel-tabset
### Code

```{r}
#| eval: false
#| echo: true

library(glmnet)

# Ridge regression (alpha = 0)
ridge_cv <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)
ridge_model <- glmnet(X_train, y_train, alpha = 0, lambda = ridge_cv$lambda.min)

# Lasso regression (alpha = 1)
lasso_cv <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
lasso_model <- glmnet(X_train, y_train, alpha = 1, lambda = lasso_cv$lambda.min)

# Elastic net (alpha between 0 and 1)
elastic_cv <- cv.glmnet(X_train, y_train, alpha = 0.5, nfolds = 10)
```

### When to Use

| Method          | Use When                                                |
|:----------------|:--------------------------------------------------------|
| **Ridge**       | Many predictors, most relevant; multicollinearity       |
| **Lasso**       | Feature selection important; many irrelevant predictors |
| **Elastic Net** | Groups of correlated predictors; best of both worlds    |
:::

## Geometric Interpretation

-   **Ridge** constraint region: circle/sphere (smooth boundary) → coefficients shrink continuously
-   **Lasso** constraint region: diamond/polytope (corners at axes) → coefficients can be exactly zero at the corners
-   The intersection of the elliptical RSS contours with the constraint region determines the solution

## Visualizing Regularization Paths

::: panel-tabset
### Code

```{r}
#| eval: false
#| echo: true

# Fit models across range of lambda values
ridge_path <- glmnet(X_train, y_train, alpha = 0)
lasso_path <- glmnet(X_train, y_train, alpha = 1)

# Plot coefficient paths
par(mfrow = c(1, 2))
plot(ridge_path, xvar = "lambda", main = "Ridge Regression Path")
plot(lasso_path, xvar = "lambda", main = "Lasso Regression Path")
```

### Interpretation

-   Each line represents one coefficient as λ changes
-   Ridge: all coefficients shrink together toward zero
-   Lasso: coefficients drop to exactly zero at different λ values
-   Cross-validation selects the optimal λ
:::

## Regularization Worked Example: Biomaterial Properties

::: panel-tabset
### Output

```{r}
#| label: reg-worked-output
#| echo: false
#| eval: true
#| fig-width: 10
#| fig-height: 5

library(glmnet)

# Simulated biomaterial data: 15 material properties predicting biocompatibility
set.seed(42)
n <- 100; p <- 15
X <- matrix(rnorm(n * p), n, p)
colnames(X) <- paste0("property_", 1:p)
y <- 3 * X[,1] - 2 * X[,3] + 1.5 * X[,7] + 0.8 * X[,12] + rnorm(n, 0, 2)

set.seed(42)
train_idx <- sample(n, 0.7 * n)
X_train <- X[train_idx, ]; y_train <- y[train_idx]
X_test  <- X[-train_idx, ]; y_test  <- y[-train_idx]

lasso_cv <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
ridge_cv <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)

# Coefficients
cat("=== Lasso Selected Coefficients (non-zero) ===\n")
lasso_coefs <- coef(lasso_cv, s = "lambda.min")
lasso_nz <- lasso_coefs[lasso_coefs[,1] != 0, , drop = FALSE]
print(round(lasso_nz, 3))

# Test RMSE comparison
lasso_pred <- predict(lasso_cv, X_test, s = "lambda.min")
ridge_pred <- predict(ridge_cv, X_test, s = "lambda.min")
ols_model  <- lm(y_train ~ X_train)
ols_pred   <- cbind(1, X_test) %*% coef(ols_model)

cat("\n=== Test Set RMSE ===\n")
cat("Lasso:", round(sqrt(mean((y_test - lasso_pred)^2)), 3), "\n")
cat("Ridge:", round(sqrt(mean((y_test - ridge_pred)^2)), 3), "\n")
cat("OLS:  ", round(sqrt(mean((y_test - ols_pred)^2)), 3), "\n")
```

### Visualization

```{r}
#| label: reg-worked-viz
#| echo: false
#| eval: true
#| fig-width: 10
#| fig-height: 4

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
plot(lasso_cv, main = "Lasso CV")
plot(ridge_cv, main = "Ridge CV")
plot(glmnet(X_train, y_train, alpha = 1), xvar = "lambda",
     main = "Lasso Coefficient Path")
abline(v = log(lasso_cv$lambda.min), lty = 2, col = "red")
legend("topright", "lambda.min", lty = 2, col = "red", cex = 0.8)
```

### Code

```{r}
#| label: reg-worked-code
#| echo: true
#| eval: false

library(glmnet)

# Simulated: 15 material properties, only 4 truly matter
set.seed(42)
n <- 100; p <- 15
X <- matrix(rnorm(n * p), n, p)
colnames(X) <- paste0("property_", 1:p)
y <- 3 * X[,1] - 2 * X[,3] + 1.5 * X[,7] + 0.8 * X[,12] + rnorm(n, 0, 2)

# Train/test split
train_idx <- sample(n, 0.7 * n)
X_train <- X[train_idx, ]; y_train <- y[train_idx]
X_test  <- X[-train_idx, ]; y_test  <- y[-train_idx]

# Cross-validated Lasso and Ridge
lasso_cv <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
ridge_cv <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)

# Extract Lasso coefficients — which properties survived?
coef(lasso_cv, s = "lambda.min")
```

### Interpretation

-   **Lasso selects the correct predictors** — properties 1, 3, 7, and 12 have non-zero coefficients, matching the true data-generating process. The remaining 11 are zeroed out.
-   **Ridge keeps all 15 predictors** — none are exactly zero, but the irrelevant ones are shrunk close to zero.
-   **Test RMSE**: Lasso and Ridge both outperform OLS on new data because regularization prevents overfitting.
-   **The CV curves** show MSE vs. log(λ). The vertical dashed lines mark lambda.min (best prediction) and lambda.1se (most regularized model within 1 SE — more conservative).
:::

------------------------------------------------------------------------
